---
title: "Project_2_GDP_Country_Data"
author: "Khandker Qaiduzzaman"
format: html
editor: visual
---

# Objective

Transform the wide-format GDP dataset into a tidy (long) format using the `tidyr` and `dplyr` packages in R. Once the dataset has been cleaned up, examine GDP patterns across nations and historical periods to see how national economies have evolved.

# Approach

### Dataset description

I'll be using a dataset of countries' GDP figures from 1960 to 2022 for this project. The dataset is available here:

<https://raw.githubusercontent.com/NafeesKhandker/Project-2-Tidy-Format-Transformation/refs/heads/main/GDP_by_Country.csv>

GDP figures for numerous nations across several decades are included in the dataset. The dataset is currently stored in a wide format where a country is represented by each row and a year by each column.

For instance, GDP values for those particular years are shown in columns like 1960, 1961, 1962, and so on.

Because the year variable is contained within column names rather than being expressed as its own column, this structure does not adhere to tidy data principles. Every variable should have its own column and every observation should have its own row in a neat dataset.

The following important variables are present in the dataset:

-   **Country Name**: Name of the country
-   **Country Code**: Country identifier
-   **1960–2022 columns**: GDP values for each year

The year columns must be transformed into a single column that represents the year and stores GDP figures in a separate column in order for the dataset to be analyzed.

### Anticipated Challenges

Managing the numerous year columns that span more than 60 years is one expected problem. During the cleaning process, these columns must be transformed into a single variable that represents the year.

Dealing with missing values for specific nations and years presents another possible difficulty. Calculations and visualizations must be done carefully because certain countries might not have GDP statistics available for the full time period.

Furthermore, GDP figures can be extremely high, therefore in order to make trend visualizations easier to read, the scale may need to be changed.

### Implementation of Data Import

```{r}
library(tidyverse) 
library(gt)

url <- "https://raw.githubusercontent.com/NafeesKhandker/Project-2-Tidy-Format-Transformation/refs/heads/main/GDP_by_Country.csv"

df <- read_csv(file = url, show_col_types = FALSE)

df |> 
  head(n = 5) |> 
  gt()
```

```{r}
library(tidyverse)
library(gt)

# URL to GDP CSV
url <- "https://raw.githubusercontent.com/NafeesKhandker/Project-2-Tidy-Format-Transformation/refs/heads/main/GDP_by_Country.csv"

# Read CSV into R
df <- read_csv(file = url, show_col_types = FALSE)

# Preview first 5 rows
df |> 
  head(5) |> 
  gt() |> 
  tab_header(
    title = "Preview of Raw GDP Data",
    subtitle = "Wide-format dataset with countries and years"
  )
```

### Reshape from Wide to Long (Tidy) Format

Convert columns `1960–2022` into a single `Year` column and a `GDP` value column. This makes the dataset tidy for analysis.

```{r}
# Reshape from wide to long (tidy) format
gdp_long <- df |> 
  pivot_longer(
    cols = 3:last_col(),  # all year columns
    names_to = "Year",
    values_to = "GDP"
  )

# Convert Year to integer and GDP to numeric
gdp_long <- gdp_long |> 
  mutate(
    Year = as.integer(Year),
    GDP = as.numeric(GDP)
  )

# Preview first 5 rows
gdp_long |> 
  head(5) |> 
  gt() |> 
  tab_header(
    title = "Tidy GDP Data (Country-Year Format)",
    subtitle = "Each row is one country in one year"
  )
```

### Normalize Variable Structure

Ensure the data types are consistent. Convert `Year` to integer and `GDP` to numeric.

```{r}
# FILTER ONLY REAL COUNTRY ECONOMIES — remove groupings/regions
gdp_long <- gdp_long |> 
  filter(
    !grepl(
      pattern = paste(
        # common grouping keywords to exclude
        "World|income|Africa|Asia|Europe|Latin America|North America|Middle East|Pacific|Caribbean|Sub-Saharan|regional|area|aggregate|total|blend|only|Not classified|OECD|Post|Late|Early|Fragile|Least|Pre|Heavily|Small",
        sep = "|"
      ),
      x = Country,
      ignore.case = TRUE
    )
  )

# Preview first 10 rows
gdp_long |> 
  filter(!is.na(GDP)) |>  # show rows with actual GDP
  head(5) |> 
  gt() |> 
  tab_header(
    title = "Normalized GDP Data for Actual Countries",
    subtitle = "Excludes regions/aggregates; non-missing GDP only"
  )
```

### Handle Missing or Inconsistent Values

Address `NA` values. For this dataset, we will keep `NA` as-is but note that these indicate missing GDP data for that year/country.

```{r}
# Count missing GDP values
gdp_long |> 
  summarise(Missing_GDP = sum(is.na(GDP)))

# Remove rows with missing GDP for analysis
gdp_long <- gdp_long |> 
  drop_na(GDP)

# Confirm no missing values remain
gdp_long |> 
  summarise(Missing_GDP = sum(is.na(GDP)))
```

```         
```

### Summary Statistics by Country

The table shows the range and average GDP for each country. Larger economies like the US or China have very high mean GDP, while smaller countries or regions have lower GDP. This helps identify countries with consistent growth versus volatile economies.

```{r}
library(scales)  # for number formatting

# Helper function to format numbers
format_gdp <- function(x) {
  ifelse(x >= 1e12, paste0(round(x/1e12, 2), " T"),
         ifelse(x >= 1e9, paste0(round(x/1e9, 2), " B"),
                ifelse(x >= 1e6, paste0(round(x/1e6, 2), " M"),
                       paste0(round(x/1e3, 2), " K"))))
}

# Summary statistics by country
gdp_summary <- gdp_long |> 
  group_by(Country) |> 
  summarise(
    Min_GDP = min(GDP),
    Max_GDP = max(GDP),
    Mean_GDP = mean(GDP),
    Median_GDP = median(GDP),
    .groups = "drop"
  ) |> 
  arrange(desc(Mean_GDP))

# Display top 10 countries
gdp_summary |> 
  gt() |> 
  fmt(
    columns = vars(Min_GDP, Max_GDP, Mean_GDP, Median_GDP),
    fns = format_gdp
  ) |> 
  tab_header(
    title = "Summary Statistics of GDP by Country",
    subtitle = "Values in USD (formatted in K, M, B)"
  ) |> 
  cols_label(
    Country = "Country",
    Min_GDP = "Min GDP",
    Max_GDP = "Max GDP",
    Mean_GDP = "Mean GDP",
    Median_GDP = "Median GDP"
  )
```

### GDP Trend Over Time (Line Plot)

The line chart highlights GDP growth over time. Economies of the top countries show upward trends, with occasional dips corresponding to economic recessions or regional crises. Normalizing to Trillions of USD makes the chart readable.

```{r}
library(ggplot2)

# Select top 10 countries by mean GDP
top10_countries <- gdp_summary |> 
  slice_head(n = 10) |> 
  pull(Country)

# Filter tidy data for top 10 countries
gdp_top10 <- gdp_long |> 
  filter(Country %in% top10_countries)

# Plot GDP trends
ggplot(gdp_top10, aes(x = Year, y = GDP, color = Country)) +
  geom_line(size = 1) +
  scale_y_continuous(labels = function(x) {
    ifelse(x >= 1e12, paste0(round(x/1e12, 1), " T"),
           ifelse(x >= 1e9, paste0(round(x/1e9, 1), " B"),
                  ifelse(x >= 1e6, paste0(round(x/1e6, 1), " M"),
                         paste0(round(x/1e3, 1), " K"))))
  }) +
  labs(
    title = "GDP Trend Over Time for Top 10 Countries",
    x = "Year",
    y = "GDP (USD)",
    color = "Country"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

### Percentage Contribution of Top 10 Countries to Total GDP

This horizontal bar chart displays the share of total GDP contributed by the top 10 countries in the dataset. Each bar represents a country, with the length proportional to its percentage of the combined GDP of all countries. The values are labeled on the bars for clarity, allowing easy comparison of economic size among the leading economies. The chart highlights which countries dominate global GDP and their relative contributions within this group.

```{r}
library(dplyr)
library(ggplot2)
library(scales)

# Calculate total GDP across all countries
total_gdp_all <- gdp_long %>%
  group_by(Country) %>%
  summarise(Total_GDP = sum(GDP, na.rm = TRUE)) %>%
  ungroup() %>%
  summarise(Global_GDP = sum(Total_GDP, na.rm = TRUE)) %>%
  pull(Global_GDP)

# Top 10 countries by total GDP
gdp_top10_summary <- gdp_long %>%
  group_by(Country) %>%
  summarise(Total_GDP = sum(GDP, na.rm = TRUE)) %>%
  ungroup() %>%
  arrange(desc(Total_GDP)) %>%
  slice_head(n = 10) %>%
  mutate(Percent_of_Total = Total_GDP / total_gdp_all * 100)

# Bar chart
ggplot(gdp_top10_summary, aes(x = reorder(Country, Percent_of_Total), y = Percent_of_Total, fill = Country)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(round(Percent_of_Total, 1), "%")), 
            hjust = -0.1, size = 4) +
  coord_flip() +  # horizontal bars for readability
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "Percentage Contribution of Top 10 Countries to Global GDP",
    x = "Country",
    y = "Percentage of Global GDP (%)"
  ) +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
  )
```

# Conclusion

The bar chart highlights that the United States, China, and Japan dominate the global economy among the top 10 countries, contributing a substantial share of total GDP. While the U.S. and China account for the largest portions, smaller economies like Mexico and the Netherlands have comparatively minor contributions. This visualization makes it easy to compare the relative economic weight of these leading countries and understand their impact on global economic trends. Overall, it underscores the concentration of global economic power in a few major nations.

# References

1.  OpenAI. (2026, March 8). *ChatGPT conversation with K. M. Qaiduzzaman on data analysis, GDP visualization, and R programming.* Retrieved from <https://chat.openai.com/>
