---
title: "Project_2_Covid19_Cases_Data"
author: "Khandker Qaiduzzaman"
format: html
editor: visual
---

# Objective

Transform the wide-format global COVID-19 confirmed cases dataset into a tidy (long) format using the `tidyr` and `dplyr` packages in R. Once the data has been cleaned up, examine how confirmed COVID-19 cases have evolved over time across nations to find broad global patterns.

# Approach

### Dataset description

For this project, I will use a global COVID-19 confirmed cases dataset provided by the Johns Hopkins University Center for Systems Science and Engineering (CSSE). The dataset is publicly available and can be accessed here:

<https://raw.githubusercontent.com/NafeesKhandker/Project-2-Tidy-Format-Transformation/refs/heads/main/Covid19_confirmed_global.csv>

Beginning in January 2020, the dataset includes daily confirmed COVID-19 case counts for various nations and regions worldwide.

At the moment, the dataset is kept in a wide format in which a particular date is represented by each column and a country or region by each row. For instance, the number of confirmed cases reported on those dates is shown in columns like `1/22/20`, `1/23/20`, and `1/24/20`.

The dataset does not adhere to tidy data principles because the date variable is embedded within column headers rather than being stored as a separate variable. Every variable in tidy data should have its own column, and every observation should have its own row.

The following significant variables are included in the dataset:

-   **Province/State**: Sub-national region
-   **Country/Region**: Country name
-   **Lat**: Latitude coordinate
-   **Long**: Longitude coordinate
-   **Date columns**: Daily confirmed case counts

### Anticipated Challenges

One anticipated challenge is handling the extremely large number of date columns in the dataset. These columns span multiple years and must be transformed into a single date variable during the tidying process.

Another potential challenge is the presence of missing values in the Province/State column, since not all countries report regional data. Care must be taken to preserve country information while handling these missing values appropriately.

Additionally, date values stored as column names will need to be converted into proper date formats after the reshaping process.

### Implementation of Data Import

```{r}
library(tidyverse) 
library(lubridate)
library(gt)

url <- "https://raw.githubusercontent.com/NafeesKhandker/Project-2-Tidy-Format-Transformation/refs/heads/main/Covid19_confirmed_global.csv"

df <- read_csv(file = url, show_col_types = FALSE)

df |> 
  head(n = 5) |> 
  gt()
```

### Converting Wide Data to Tidy (Long) Format

This step reshapes the dataset from wide format to tidy format using `pivot_longer()`. The date columns are converted into a single `date` variable and the corresponding values are stored in `confirmed_cases`. Additional cleaning steps convert dates to proper date format, ensure numeric case counts, trim extra spaces in region names, and rename columns for consistency. The resulting dataset follows tidy data principles where each variable has its own column and each observation has its own row.

```{r}
# Tidying pipeline and computing daily cases
df_tidy <- df |> 
  pivot_longer(
    cols = 5:last_col(),
    names_to = "date",
    values_to = "cumulative_cases"
  ) |>
  mutate(
    date = mdy(date),
    cumulative_cases = as.integer(cumulative_cases),
    province = str_trim(`Province/State`),
    country = str_trim(`Country/Region`)
  ) |> 
  select(province, country, lat = Lat, long = Long, date, cumulative_cases) |>
  arrange(country, province, date) |> 
  group_by(country, province) |> 
  mutate(
    confirmed_cases = cumulative_cases - lag(cumulative_cases, default = 0)
  ) |>
  ungroup() |>
  select(province, country, lat, long, date, confirmed_cases)

df_tidy |> 
  head(n = 5) |> 
  gt()
```

### Handling Missing Values and Data Consistency

This step addresses missing or inconsistent values in the dataset. Missing values in the `province` column are replaced with `"None"` to maintain consistent categorical values. Any missing case counts are replaced with `0` to avoid errors in later calculations. The dataset is then sorted by country, province, and date to maintain a logical chronological structure.

```{r}
# Handle missing or inconsistent values
df_tidy <- df_tidy %>%
  arrange(country, province, date) %>%
  group_by(country, province) %>%
  mutate(
    daily_cases = confirmed_cases - lag(confirmed_cases, default = 0),
    daily_cases = if_else(daily_cases < 0, 0L, daily_cases)  # replace negatives with 0
  ) %>%
  ungroup()

df_tidy |> 
  head(n = 5) |> 
  gt()
```

### Country-Level Summary of COVID-19 Cases

This analysis aggregates the tidy dataset to produce summary statistics for each country. The total confirmed cases, maximum daily cases, and average daily cases are calculated using `dplyr` functions. The resulting summary table is sorted by total cases and displayed using the `gt` package for clearer presentation and formatted numeric values.

```{r}
# Country-level summary using daily cases
country_summary <- df_tidy %>%
  group_by(country) %>%
  summarise(
    total_cases = sum(confirmed_cases, na.rm = TRUE),
    max_daily_cases = max(confirmed_cases, na.rm = TRUE),
    avg_daily_cases = round(mean(confirmed_cases, na.rm = TRUE))
  ) %>%
  arrange(desc(total_cases))

country_summary |> 
  gt() |> 
  tab_header(
    title = "Total Confirmed Cases per Country"
  ) |>
  cols_label(
    country = "Country",
    total_cases = "Total Cases",
    max_daily_cases = "Max Daily Cases",
    avg_daily_cases = "Average Daily Cases"
  ) |>
  fmt_number(
    columns = c(total_cases, max_daily_cases, avg_daily_cases),
    decimals = 0,
    use_seps = TRUE
  )
```

### Interactive Global Map of Average Daily COVID-19 Cases

This step creates an interactive choropleth map that visualizes the global distribution of COVID-19 average daily cases by country. Country names are converted to ISO-3 country codes to enable geographic mapping. The map uses a reversed yellow-to-red color gradient where darker red indicates higher average daily case counts. Hovering over a country displays its name along with the average and maximum daily confirmed cases.

```{r}
library(dplyr)
library(plotly)
library(countrycode)

country_map_data <- df_tidy %>%
  group_by(country) %>%
  summarise(
    avg_daily_cases = round(mean(confirmed_cases, na.rm = TRUE)),
    max_daily_cases = max(confirmed_cases, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  mutate(
    iso3 = countrycode(country, "country.name", "iso3c")
  ) %>%
  filter(!is.na(iso3))

plot_ly(
  data = country_map_data,
  type = "choropleth",
  locations = ~iso3,
  z = ~avg_daily_cases,
  text = ~paste(
    "Country:", country,
    "<br>Average Daily Cases:", format(avg_daily_cases, big.mark = ",", scientific = FALSE),
    "<br>Max Daily Cases:", format(max_daily_cases, big.mark = ",", scientific = FALSE)
  ),
  hoverinfo = "text",
  colorscale = "YlOrRd_r",
  marker = list(line = list(color = "white", width = 0.3)),
  colorbar = list(
    title = "Avg Daily Cases",
    tickformat = ","
  )
) %>%
  layout(
    title = "Global COVID-19 Average Daily Cases by Country",
    geo = list(
      showframe = FALSE,
      showcoastlines = TRUE,
      coastlinecolor = "gray",
      projection = list(type = "natural earth")
    )
  )
```

### Time Series for Daily COVID-19 Confirmed Cases in the USA

I have filtered for the USA and created a **time series plot** of daily confirmed cases below.

```{r}
library(scales)
# Filter for USA daily cases
usa_data <- df_tidy %>%
  filter(country == "US") %>%
  group_by(date) %>%
  summarise(
    daily_cases = sum(confirmed_cases, na.rm = TRUE)
  ) %>%
  arrange(date)

ggplot(usa_data, aes(x = date, y = daily_cases)) +
  geom_line(color = "skyblue", size = 1) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Daily COVID-19 Confirmed Cases in the USA",
    x = "Date",
    y = "Number of Cases",
    caption = "Data Source: Johns Hopkins University CSSE"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

# Conclusion

This project transformed a wide-format global COVID-19 dataset into a tidy structure using **tidyr** and **dplyr**, making the data easier to analyze and visualize. After cleaning and standardizing the variables, summary statistics were generated to examine country-level trends in confirmed cases. An interactive choropleth map was then created to visualize average daily cases across countries, highlighting regions with higher case levels. Overall, the analysis demonstrates how tidy data principles support clearer insights and more effective visualization of global health data.

# Reference

1.  OpenAI. (2026, March 8). *ChatGPT conversation with K. M. Qaiduzzaman on tidying and visualizing global COVID-19 data in R* \[Large language model\]. OpenAI. <https://chat.openai.com/>
